High Resolution Identification of Protein-DNA Binding Events and Quality Control for ChIP-exo data

### 2015-11-19 rmarkdown/chip_exo_paper.Rnw
Added ChIPnexus paper to bibliography and added section of biases comparison of ChIPexo for eukaryotic genomes

DNA libraries generated by the ChIP-exo protocol seem to be less
complex than the libraries generated by ChIP-Seq
\cite{exo_review}. Hence, most of current guidelines \cite{encode_qc}
may not be applicable on ChIP-exo, additionally to our knowledge there
are not established quality control pipelines for ChIP-exo. To address
this challenge, we suggest a series of quality control visualizations
to understand which biases are present in ChIP-exo data. Related to
quality control, Previous ChIP-exo analysis used ChIP-Seq samples to
compare the resolution between experiments (\cite{exo1}, \cite{exo2},
\cite{exoillumina}). In \cite{carroll.qc}, Carroll et. al. studied the
use of the Strand-Cross Correlation (SCC) \cite{strandcc}. and showed
that by filtering blacklisted regions the estimation of the SCC is
improved. However, this method requires to know blacklisted regions in
advance which may not be available. In our pipeline we propose two
out-the-shelf metrics equivalent to RSC and NSC for the estimation of
the signal-to-noise ratio for a ChIP-exo sample.

@@

To have a first view on the data, we calculated a series of the most
commonly used quality indicators for ChIP-Seq data, which are listed
in table \ref{tab:qcbase}. The PBC, NSC and RSC were calculated as
defined in \cite{encode_qc} and the SSD was calculated as defined in
\cite{htseq}. Additionally, we calculated the FSR to compare the
number of fragments in each strand. A few conclusions that we may get
by observing the PBC in table \ref{tab:qcbase} are all the samples
have a quite low library complexity, hence it would be suggested to
repeat the experiment.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width = .9\textwidth]{../figs/for_paper/coverage_diagram.pdf}
  \caption{Diagram of the ChIP-exo Quality control pipeline(this is a
    place-holder). The genome is partitioned into several regions by
    removing the empty segments and for each regions we calculate a
    collection of qc indicators.}
  \label{fig:qcdiagram}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
  \includegraphics[width = .8\textwidth,page =3]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata.pdf}
  \includegraphics[width = .8\textwidth,page =4]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata_sameRegion.pdf}
  \caption{Using the mouse-FoxA1 experiment from \cite{exoillumina}:
    A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$, B) Boxplots
    of the $\mbox{local-NSC}$ stratified by nr. of reads mapped to
    only one position in the regions for the ChIP-exo experiments
    divided by condition and biological replicate and C) Boxplots of
    the a set of ERR336942 regions, stratified as in (B) with the
    $\mbox{local-NSC}$ calculated with the different biological
    replicates.}
  \label{fig:enrich}
\end{figure}

@@

In order to asses the strand imbalance we create the following
visualization in figure \ref{fig:strand}. A) presents the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strand or only one. Roughly speaking in a lower quality dataset there
is going to be a higher amount of single stranded-regions. To evaluate
this statement, for each replicate of the FoxA1 dataset from
\cite{exoillumina}, we divided the regions formed by at least
\Sexpr{de} fragments into into the ones that overlap with any of the
sample's peaks. The p.value for ERR36935 is \Sexpr{pval[[1]]}, for
ERR336942 is \Sexpr{pval[[2]]} and for ERR336956 is \Sexpr{pval[[3]]}.

\begin{figure}[h!]
  \centering  
  \includegraphics[width = .8\textwidth,page = 3]{../figs/Carroll_mice_for_paper/Strand_imbalance.pdf} 
  \caption{Strand imbalance QC plots for the same data as in figure \ref{fig:enrich}.}
  \label{fig:strand}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width = .46\textwidth,page = 3 ]{../figs/for_paper/ChIPseqPET_ChIPexo_tagCount_comparison.pdf}
  \includegraphics[width = .46\textwidth]{../figs/for_paper/forward_strand_ratio_comp_old.pdf}
  \caption{ A) ChIP signal in ChIP-exo was linearly related to that in
    PET ChIP-Seq data in the region with high ChIP tag counts. In
    contrast, there were clear differences in their background
    distribution, where several ChIP-exo background regions were
    empty. B) In ChIP-exo data, strands of reads were significantly
    less balanced in the regions with potential binding sites compared
    to SET ChIP-Seq data. }
  \label{fig:comp}
\end{figure}

@@

>>>>

DNA libraries generated by the ChIP-exo protocol seem to be less
complex than the libraries generated by ChIP-Seq
\cite{exo_review}. Hence, most of current guidelines \cite{encode_qc}
may not be applicable on ChIP-exo, additionally to our knowledge there
are not established quality control pipelines for ChIP-exo. To address
this challenge, we suggest a collection of quality control
visualizations to understand which biases are present in ChIP-exo
data. Related to quality control, Previous ChIP-exo analysis used
ChIP-Seq samples to compare the resolution between experiments
(\cite{exo1}, \cite{exo2}, \cite{exoillumina}). In \cite{carroll.qc},
Carroll et. al. studied the use of the Strand-Cross Correlation (SCC)
\cite{strandcc}. and showed that by filtering blacklisted regions the
estimation of the SCC is improved. However, this method requires to
know blacklisted regions in advance which may not be available. In our
pipeline we propose two out-the-shelf metrics equivalent to RSC and
NSC for the estimation of the signal-to-noise ratio for a ChIP-exo
sample.

@@

To have a first view on the data, we calculated the most commonly used
quality indicators for ChIP-Seq data, which are listed in table
\ref{tab:qcbase}. The PBC, NSC and RSC were calculated as defined in
\cite{encode_qc} and the SSD was calculated as defined in
\cite{htseq}. Additionally, we calculated the FSR to compare the
number of fragments in each strand. A few conclusions that we may get
by observing the PBC in table \ref{tab:qcbase} are all the samples
have a quite low library complexity, hence it would be suggested to
repeat the experiment.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width = .9\textwidth]{../figs/for_paper/coverage_diagram.pdf}
  \caption{Diagram of the ChIP-exo Quality control pipeline [this is a
    place-holder]. The genome is partitioned into several regions by
    removing the empty segments and for each regions we calculate a
    collection of qc indicators.}
  \label{fig:qcdiagram}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
  \includegraphics[width = .8\textwidth,page =3]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata.pdf}
  \includegraphics[width = .8\textwidth,page =4]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata_sameRegion.pdf}
  \caption{Using the mouse-FoxA1 experiment from \cite{exoillumina}:
    A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$, in general
    we can see a slight separation into two strong arms, one
    corresponds to low $\mbox{ARC}$ and varying $\mbox{URCR}$, and for
    the other $\mbox{URCR}$ decreases as $\mbox{ARC}$ increases B)
    Boxplots of the $\mbox{local-NSC}$ stratified by nr. of reads
    mapped to only one position in the regions for the ChIP-exo
    experiments divided by condition and biological replicate and C)
    Boxplots of the a set of ERR336942 regions, stratified as in (B)
    with the $\mbox{local-NSC}$ calculated with the different
    biological replicates.}
  \label{fig:enrich}
\end{figure}

@@

In order to asses the strand imbalance we create the following
visualization in figure \ref{fig:strand}. A) presents the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strand or only one. Roughly speaking in a lower quality dataset there
is going to be a higher amount of single stranded-regions. To evaluate
this statement, for each replicate of the FoxA1 dataset from
\cite{exoillumina}, we divided the regions formed by at least
\Sexpr{de} fragments into into the ones that overlap with any of the
sample's peaks. The p.value for ERR36935 is \Sexpr{pval[[1]]}, for
ERR336942 is \Sexpr{pval[[2]]} and for ERR336956 is
\Sexpr{pval[[3]]}. This shows that only the middle sample from figure
\ref{fig:strand} is the least imbalance sample from the the three,
while the other two still show a higher imbalance.

\begin{figure}[h!]
  \centering  
  \includegraphics[width = .8\textwidth,page = 3]{../figs/Carroll_mice_for_paper/Strand_imbalance.pdf} 
  \caption{Strand imbalance QC plots for the same data as in figure
    \ref{fig:enrich}. A) FSR distribution quantiles as the lower depth
    regions are being filtered out. B) Stacked histogram with the
    proportion of regions that are formed by two strands or only one.}
  \label{fig:strand}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width = .46\textwidth,page = 3 ]{../figs/for_paper/ChIPseqPET_ChIPexo_tagCount_comparison.pdf}
  \includegraphics[width = .46\textwidth]{../figs/for_paper/forward_strand_ratio_comp_old.pdf}
  \caption{ A) ChIP signal in ChIP-exo was linearly related to that in
    PET ChIP-Seq data in the region with high ChIP tag counts. In
    contrast, there were clear differences in their background
    distribution, where several ChIP-exo background regions were
    empty. B) In ChIP-exo data, strands of reads were significantly
    less balanced in the regions with potential binding sites compared
    to SET ChIP-Seq data. [C and D correspond to fig4 from manuscript]
    C) ChIP tag counts increase linearly as mappability scores
    increases. D) ChIP tag counts increase linearly as GC content
    score increases when GC content is less than 0.6 and then ChIP tag
    counts decrease as GC content increases. }
  \label{fig:comp}
\end{figure}

\subsubsection{ChIP-exo data from eukaryotic genome shows clear sequence biases}
\label{sec:eukaryotic}

We evaluated ChIP-exo data for CTCF factor from human genome
\cite{exo1} to investigate issues specific to eukaryotic genomes for
binding sites identification. Figures \ref{fig:comp}C and
\ref{fig:comp}D display the bin-level average read counts against
mappability and GC content. Each data point is obtained by averaging
the read counts across bins with the same mappability of GC
content. These results indicate that binding site identification in
ChIP-exo sample might also benefic from the use of methods that take
into account apparent sequence biases such as mappability and GC
content.

@@

\subsection{dPeak outperforms compeing methods in discovering closely
  space binding event from ChIP-exo and ChIP-Seq data}
\label{sec:dpeak_comp}

\emph{I plan to update this analysis, since the dpeak's new strategy
  seems to decrease resolution. Also, Apex was updated into MACE, so I
  think it could be better to add that method instead}

### 2015-11-19 rmarkdown/chip_exo_paper.tex
Added ChIPnexus paper to bibliography and added section of biases comparison of ChIPexo for eukaryotic genomes

In order to asses the strand imbalance we create the following
visualization in figure \ref{fig:strand}. A) presents the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strand or only one. Roughly speaking in a lower quality dataset there
is going to be a higher amount of single stranded-regions. To evaluate
this statement, for each replicate of the FoxA1 dataset from
\cite{exoillumina}, we divided the regions considered by the QC
pipeline into two sets by reviewing if they overlap any of the
sample's peaks and then calculate the Kolmogorov-Smirnov test.

\begin{figure}[h!]
  \centering  
  \includegraphics[width = .8\textwidth,page = 3]{../figs/Carroll_mice_for_paper/Strand_imbalance.pdf} 
  \caption{Strand imbalance QC plots for the same data as in figure \ref{fig:enrich}.}
  \label{fig:strand}
\end{figure}

>>>>

In order to asses the strand imbalance we create the following
visualization in figure \ref{fig:strand}. A) presents the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strand or only one. Roughly speaking in a lower quality dataset there
is going to be a higher amount of single stranded-regions. To evaluate
this statement, for each replicate of the FoxA1 dataset from
\cite{exoillumina}, we divided the regions formed by at least
100 fragments into into the ones that overlap with any of the
sample's peaks. The p.value for ERR36935 is 0.1039, for
ERR336942 is 0.0064 and for ERR336956 is
0.0871. This shows that only the middle sample from figure
\ref{fig:strand} is the least imbalance sample from the the three,
while the other two still show a higher imbalance.

\begin{figure}[h!]
  \centering  
  \includegraphics[width = .8\textwidth,page = 3]{../figs/Carroll_mice_for_paper/Strand_imbalance.pdf} 
  \caption{Strand imbalance QC plots for the same data as in figure
    \ref{fig:enrich}. A) FSR distribution quantiles as the lower depth
    regions are being filtered out. B) Stacked histogram with the
    proportion of regions that are formed by two strands or only one.}
  \label{fig:strand}
\end{figure}
